Arrows for Parallel Computations

### 2017-12-07 paper/content/arrows.tex
reworked some stuff in the arrows chapter

@@

\subsection{Arrows}
\label{sec:arrows}
Arrows were introduced by \citet{HughesArrows} as a general interface for computation. An Arrow |arr a b| represents  a computation that converts an input |a| to an output |b|. This is defined in the |Arrow| type class shown in Fig.~\ref{fig:ArrowDefinition}.
%
To lift an ordinary function to an Arrow, |arr| is used, analogous to the monadic |return|. Similarly, the composition operator |>>>| is analogous to the monadic composition |>>=| and combines two arrows |arr a b| and |arr b c| by \enquote{wiring} the outputs of the first to the inputs to the second to get a new arrow |arr a c|. Lastly, the |first| operator takes the input Arrow |arr a b| and converts it into an arrow on pairs |arr (a, c) (b, c)| that leaves the second argument untouched. It allows us to to save input across arrows. Figure~\ref{fig:arrows-viz} shows a graphical representation of these basic Arrow combinators.
The most prominent instances of this interface are regular functions |(->)|
and the Kleisli type (Fig.~\ref{fig:ArrowDefinition}), which wraps monadic functions, e.g.  |a -> m b| (with |m| being a Monad).

@@

The more restrictive interface of Arrows allows for more elaborate composition and transformation combinators---a Monad can be \emph{anything}, an Arrow is a process of doing something, a \emph{computation}. This is exactly one of the key challenges in parallel computing.

>>>>

\subsection{Arrows}
\label{sec:arrows}
Arrows were introduced by \citet{HughesArrows} as a general interface for computation and a less restrictive generalisation of Monads. \citeauthor{HughesArrows} motivates the broader interface of Arrows with the example of a parser with added static meta-information that can not satisfy the monadic bind operator |(>>=) :: m a -> (a -> m b) -> m b| (with |m| being a Monad).

An Arrow |arr a b| represents a computation that converts an input |a| to an output |b|. This is defined in the |Arrow| type class shown in Fig.~\ref{fig:ArrowDefinition}.
%
To lift an ordinary function to an Arrow, |arr| is used, analogous to the monadic |return|. Similarly, the composition operator |>>>| is analogous to the monadic composition |>>=| and combines two arrows |arr a b| and |arr b c| by \enquote{wiring} the outputs of the first to the inputs to the second to get a new arrow |arr a c|. Lastly, the |first| operator takes the input Arrow |arr a b| and converts it into an arrow on pairs |arr (a, c) (b, c)| that leaves the second argument untouched. It allows us to to save input across arrows. Figure~\ref{fig:arrows-viz} shows a graphical representation of these basic Arrow combinators.
The most prominent instances of this interface are regular functions |(->)|
and the Kleisli type (Fig.~\ref{fig:ArrowDefinition}), which wraps monadic functions, e.g.  |a -> m b|.

Note that with the Kleisli type we can define an alternative to |(>>=)| with the signature |m a -> Kleisli m a b -> m b|. The difference between the definitions of Arrows and Monads becomes more clear when we compare this definition with the Arrow compose operator |(>>>)|: Monads are types that allow composition of computations, while Arrows \emph{are} computations.

@@

@@

### 2017-12-07 paper/content/conclusion.tex
reworked some stuff in the arrows chapter

\section{Conclusion}
\label{sec:conclusion}
Arrows are a generic concept that allows for powerful composition
combinators. To our knowledge we are first to represent
\emph{parallel} computation with Arrows.
% \olcomment{that strange arrows-based robot interaction paper from 1993 or so! clearly discuss in related work} \done
%
Arrows turn out to be a useful tool for composing parallel
programs. We do not have to introduce new monadic types that wrap the
computation. Instead, we use Arrows in the same manner one uses sequential pure functions. 
%
This work features multiple parallel backends: the already available parallel Haskell flavours. Parallel Arrows, as presented here, feature an implementation of the |ArrowParallel| type class for GpH Haskell, |Par| Monad, and Eden. With our approach parallel programs can be ported across these flavours with little to no effort. It is quite straightforward to add further backends. 
%
%
Performance-wise, Parallel Arrows are on par with existing parallel Haskells, as they do not introduce any notable overhead in our benchmarks.
%
The benefit is, however, the greatly increased portability of parallel programs.

>>>>

\section{Conclusion}
\label{sec:conclusion}
Arrows are a generic concept that allows for powerful composition
combinators. To our knowledge we are first to represent
\emph{parallel} computation with Arrows.
% \olcomment{that strange arrows-based robot interaction paper from 1993 or so! clearly discuss in related work} \done
%
Arrows turn out to be a useful tool for composing parallel
programs. We do not have to introduce new monadic types that wrap the
computation. Instead, we use Arrows in the same manner one uses sequential pure functions. Even though we make heavy use of the |ArrowApply| typeclass - which only Arrows that are equivalent to Monads can satisfy - in this paper, we believe that the general idea of an Arrow, the process of doing something - a \emph{computation}, is the perfect fit to use in a library for parallel computation as it allows users to think about parallel programs more as a set of composed computations - like a data-flow graph. Furthermore, our approach allows for a more natural paralellization of any Arrow type (especially regular functions (|->|)), when compared to Monads as we do not have to change anything in order to run them in parallel.
%
This work features multiple parallel backends: the already available parallel Haskell flavours. Parallel Arrows, as presented here, feature an implementation of the |ArrowParallel| type class for GpH Haskell, |Par| Monad, and Eden. With our approach parallel programs can be ported across these flavours with little to no effort. It is quite straightforward to add further backends. 
%
%
Performance-wise, Parallel Arrows are on par with existing parallel Haskells, as they only introduce minor overhead in some of our benchmarks.
%
The benefit is, however, the greatly increased portability of parallel programs.
